---
title: "How does R generate random numbers?"
output: html_document
---

```{r, echo=FALSE}
download.file("http://www.openintro.org/stat/data/kobe.RData", destfile = "kobe.RData")
load("kobe.RData")
outcomes <- c("heads", "tails")
```

Computers can't actually generate pure random numbers.  Rather, they generate _pseudorandom_ numbers: they are _statistically indistinguishable_ from true randomness but are generated by an entirely _deterministic_ causal process.


### Seed Values
The pseudorandom number generator works by taking in a _seed_ value, which is a combination of values that we assume has no distinguishable pattern.  Example:  a seed value can be some numerical combination of

* Time of day
* Data
* Amount of space left on the hardrive
* Voltage measurements

The generator takes that seed value and then generates numbers that "look" random.  The catch: if you give the random number generator the same seed value, it gives the same pseudorandom values.

### Example
Say a HW question asks you "Say you flip a coin 1000 times, how many heads do you observe?"  You run the simulation and get:
```{r}
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
```

Say you run the simulation again.  Chances are you won't get the same results.  
```{r}
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
```

When reporting in your HW's what the answer is, you'll have to change it every time!


### Setting the same seed value
There is a solution.  You can get "replicable" pseudorandom results by manually setting the seed number.  If you don't manually set it, R uses the crazy scheme from above.

```{r}
set.seed(76)
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
set.seed(76)
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
```


### Different seed values yield different results
But if you set a different seed value, you get different results
```{r}
set.seed(76)
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
set.seed(79)
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
```

and then subsequent simulations _without_ resetting the seed yields different values

```{r}
sim_fair_coin <- sample(outcomes, size = 1000, replace = TRUE)
table(sim_fair_coin)
```

### Moral of the story
To get "replicable" random values, at the beginning of your .Rmd file, set the seed value to your favorite number just once.  

