---
title: "Histograms"
author: "Albert Y. Kim"
date: "September 25, 2015"
output: html_document
---

* A distribution indicates the frequency with which values occur.
* A histogram is a visualization of the distribution.

We load the `mosaic` and `openintro` packages, load the `county` data set, and plot this histogram for `income`.

```{r, message=FALSE}
library(mosaic)
library(openintro)
data(county)
histogram(~income, data = county)
```

What is density? Instead let's change the y-axis to reflect counts by adding the argument `type="count"`. We see the histogram has the same shape; just different values on the y-axis now.  

```{r}
histogram(~income, data = county, type="count")
```

What this is saying that about

* about 200 counties (out of 3143) have per capita income between about 12K and 16K.
* about 400 counties have per capita income between about 27K and 30K.

We can also change the y-axis to reflect percentages of the data set by adding the argument `type="percent"`. Once again, we see the histogram has the same shape.

```{r}
histogram(~income, data = county, type="percent")
```

The default `type` for `histogram` is density. Using the density values, the areas of the all the boxes sum to 1.  

```{r}
histogram(~income, data = county)
```