---
title: "Alcohol Consumption and Building QQ-Plots"
date: "`r Sys.Date()`"
output: html_document
---

We need two packages first:

1. `dplyr`: allows for fast, consistent, and convenient tools for working with data frame like objects
2. `RCurl`: allows R to to compose general HTTP requests and provides convenient functions to fetch URIs, get & post forms, etc. and process the results returned by the Web server.

```{r, echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(RCurl))
webpage <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/alcohol-consumption/drinks.csv"
raw.data <- getURL(webpage)
booze <- read.csv(text = raw.data, header=TRUE)
```

We downloaded the data used in the article [Where Do People Drink The Most Beer, Wine And Spirits?](http://fivethirtyeight.com/datalab/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/) from the [data repository](https://github.com/fivethirtyeight/data/tree/master/alcohol-consumption) and look at the top 5 drinking countries in terms of:

Hard liquor
```{r, echo=FALSE}
arrange(booze, desc(spirit_servings)) %>% 
  select(country, spirit_servings) %>% 
  head(n=5)
```

Beer
```{r, echo=FALSE}
arrange(booze, desc(beer_servings)) %>% 
  select(country, beer_servings) %>% 
  head(n=5)
```

Wine
```{r, echo=FALSE}
arrange(booze, desc(wine_servings)) %>% 
  select(country, wine_servings) %>% 
  head(n=5)
```

Total Alcohol
```{r, echo=FALSE}
arrange(booze, desc(total_litres_of_pure_alcohol)) %>% 
  select(country, total_litres_of_pure_alcohol) %>% 
  head(n=5)
```


 
```{r, fig.width=10, echo=FALSE}
alcohol <- booze$total_litres_of_pure_alcohol
xbar <- round(mean(alcohol), 3)
s <- round(sd(alcohol), 3)
```

We're going to focus on annual alcohol consumption in litres.  The observed sample mean and sample standard deviation are $\overline{x}$ =  `r xbar` and $s$ = `r s`.  In the plots below, on the left we have the observed histogram of the data, the right is simulated normal data with the same mean $\overline{x}$ and standard deviation $s$ as the observed sample.  The data on the left does not look anything like the data on the right i.e. not normal. 
```{r, fig.width=10, echo=FALSE}
par(mfrow=c(1, 2))
hist(alcohol, xlab="litres of pure alcohol", ylab="# of counties", 
     main="Observed Annual Alcohol Consumption")
legend("topright", bty='n', legend=c(paste("mean =", xbar), paste("sd =", s)))
abline(v=xbar, col="red", lwd=2)
theor.hist <- hist(rnorm(10^6, mean=xbar, sd=s), plot=FALSE, breaks=8)
theor.hist$counts <- theor.hist$density * length(alcohol)
plot(theor.hist, xlab="litres of pure alcohol", ylab="# of counties",
     main="Theoretically Normal Data")
legend("topright", bty='n', legend=c(paste("mean = ", xbar), paste("sd = ", s)))
abline(v=xbar, col="red", lwd=2)
```

We compare the observed quantiles with what the quantiles would look like for theoretically normal data.  Both plots have the same number of points.  
```{r, fig.width=10, echo=FALSE}
# A nice function that allows you to plot all
par(mfrow=c(1,2))
plot(ecdf(alcohol),
     xlab="quantiles of litres of pure alcohol", 
     ylab="proportion less than quantile", main="Observed Quantiles")

plot(ecdf(rnorm(length(alcohol), mean=xbar, sd=s)), 
     xlab="quantiles of litres of pure alcohol", 
     ylab="proportion less than quantile", main="Theoretical Normal Quantiles")
```


For both the plots above, we take the values on the x-axis and compare them on the same plot; this is a QQ-plot:
```{r, fig.width=5, echo=FALSE}
epsilon <- 0.0001
p <- seq(epsilon, 1-epsilon, length=length(alcohol))
par(mfrow=c(1, 1))
plot(qnorm(p, mean=xbar, sd=s), quantile(alcohol, probs = p), pch=19, 
     xlab="Theoretical Quantiles", ylab="Sample Quantiles",
     main="Normal Q-Q Plot")
abline(c(0, 1))
```

Given the lack of fit, we can confidently say the observed data is not normal.  
